---
title: "BMI206 Project Part 1"
author: "Wanjun Gu, Ashvin Ravi, Reet Mishra"
date: "2023-10-24"
output: html_document
---

## Data acquisition

Paper Title: **Global reference mapping of human transcription factor footprints**
Paper reference: Vierstra, J. et al. Global reference mapping of human transcription factor footprints. Nature 583, 729–736 (2020)

Data is downloaded from: https://static-content.springer.com/esm/art%3A10.1038%2Fs41586-020-2528-x/MediaObjects/41586_2020_2528_MOESM4_ESM.xlsx

## Load libraries
```{r}

suppressPackageStartupMessages({
  library(dplyr)
  library(data.table)
  library(readxl)
})
```

## Load data
```{r}
cluster = read_excel("supp table 2.xlsx", sheet = "Archetype clusters")
motifs = read_excel("supp table 2.xlsx", sheet = "Motifs")

d = left_join(cluster, motifs, by = "Cluster_ID")
```
## Summary of the dataset
```{r}
glimpse(d)
```

## Column description

| Column  descriptions                                         |
| ------------------------------------------------------------ |
| Cluster_ID: Motif cluster  number                            |
| Name: Cluster name (motif, DBD  or individual TF gene names) |
| DBD: DNA-binding domain family                               |
| Seed_motif: Motif to which  other cluster members are aligned |
| Total_width: Total with of all  aligned motifs within cluster |
| Consensus_left: Left boundary  of archetype motif            |
| Consensus_right: Right  boundary of archetype motif          |
| Cluster_size: Number of motifs  in cluster                   |
| Motif: Motif model name                                      |
| Database: Source of motif  model                             |
| Consensus: Model consensus  sequence                         |
| Relative_orientation: Relative  orientation of motof model to cluster seed motif |
| Width: Motif model width                                     |
| Left_offset: Offset relative  to left coordinate of full cluster width |
| Right_offset: Offset relative  to right coordinate of full cluster width |


```{r}
summary(d)
```

In this dataset, there are **2,174** motifs grouped into **286** clusters. For each motif/cluster, there are measurements of motif total width (*numeric*), offset relative to left/right coordinate of full cluster width (*numeric*), as well as their relative_orientations (*cetagorical/binary*). Here we showed the distribution of the length of Motifs


## Distribution of the length of Motifs
```{r}
hist(d$Width, main = "Distributions of the length of Motifs", 
     xlab = "Length", breaks = 25)
```





